---
title: "adjusting_opacity"
output:
  format: 
    latex:
      
params:
  eval_models: false
  
knitr:
  opts_chunk: 
    cache_comments: false
    crop: true
    out-width: NA
    out-height: NA
    
execute: 
  echo: false
  warning: false
  message: false
  include: false      
---

```{r}
#| label: setup

# load packages

library(tidyverse)
library(MASS)
library(emmeans)
library(scales)
library(buildmer)
library(lme4)
library(kableExtra)
library(afex)
library(papaja)
library(broom.mixed)
library(insight)
library(qwraps2)
library(lmerTest)
library(tinylabels)
library(ggdist)
library(ggpubr)
library(geomtextpath)
library(conflicted)

# fix conflicts

conflicts_prefer(dplyr::select(), dplyr::filter(), lme4::lmer())
```

```{r}
#| label: load-data

exp1_anon <- read_csv("../data/data_E1_uniform")
```



# Abstract {#abstract-adjusting-opacity}

# Preface: Learning From an Early Pilot Study {#pilot-study}

# Introduction {#introduction-adjusting-opacity}

## Overview {#overview-adjusting-opacity}

# Related Work {#related-work-adjusting-opacity}

## Transparency, Contrast, Opacity, and Formal Definitions

## Effects of Point Opacity on Correlation Estimation

# Experiment 1: Uniform Opacity Adjustments

## Introduction {#introduction-adjusting-opacity-e1}

## Methods {#methods-adjusting-opacity-e1}

## Analysis {#analysis-adjusting-opacity-e1}

## Discussion {#discussion-adjusting-opacity-e1}

# Experiment 2: Spatially-Dependent Opacity Adjustments

## Introduction {#introduction-adjusting-opacity-e2}

## Methods {#methods-adjusting-opacity-e2}

## Analysis {#analysis-adjusting-opacity-e2}

## Discussion {#discussion-adjusting-opacity-e2}

# General Discussion {#general-discussion-adjusting-opacity}
