---
title: "adjusting_size"
output:
  format: 
    latex:
params:
  eval_models: false
  
knitr:
  opts_chunk: 
    cache_comments: false
    crop: true
    out-width: NA
    out-height: NA
    cache.path: "../thesis_cache/"
    
execute: 
  echo: false
  warning: false
  message: false
  include: false     
---

```{r}
#| label: setup

# load packages

library(tidyverse)
library(MASS)
library(emmeans)
library(scales)
library(buildmer)
library(lme4)
library(kableExtra)
library(afex)
library(papaja)
library(broom.mixed)
library(insight)
library(qwraps2)
library(lmerTest)
library(tinylabels)
library(ggdist)
library(ggpubr)
library(geomtextpath)
library(conflicted)
library(formattable)
library(effectsize)

# fix conflicts

conflicts_prefer(dplyr::select(), dplyr::filter(), lme4::lmer())

# define plotting labels now

labels_e3 <- c(A = "Non-Linear\nDecay",
               B = "Linear\nDecay",
               C = "Inverted\nDecay",
               D = "Standard\nSize")

labels_e3_regex <- c(
  "\\bA\\b" = "Non-Linear Decay",
  "\\bB\\b" = "Linear Decay",
  "\\bC\\b" = "Inverted Decay",
  "\\bD\\b" = "Standard Size"
)
```

```{r}
#| label: load-data

exp3_anon <- read_csv("../data/exp_3_data.csv") %>%
  mutate("expName" = recode(expName,
                            "exp_size_only" = "E3_size_adjustments"))

source("../shared_functions.R")
```

```{r}
#| label: retrieve-cached-models-chap4

if (!params$eval_models){ lazyload_cache_dir("../thesis_cache/") }
```

```{r}
#| label: wrangle-data-chap5

# NB: With the exception of anonymisation, data are provided as-is from pavlovia.
# Wrangling functions must be run first to make the dataset usable.

wrangle <- function(anon_file) {
  
  # first do literacy
  
    literacy <- anon_file %>%
    filter(!is.na(q1_slider.response)) %>%
    rowwise() %>%
    mutate(literacy = sum(c(q1_slider.response, 
                            q2_slider.response, 
                            q3_slider.response, 
                            q4_slider.response, 
                            q5_slider.response))) %>%
    select(participant,
           literacy)
    
    # extract and process visual threshold testing
  
  visual_thresholds <- anon_file %>%
    filter(!is.na(VT_with_labels)) %>%
    select(c("VT_with_labels", "participant", "VT_textbox2.text")) %>%
    mutate(VT_answer = str_replace(VT_with_labels, pattern = "vis_threshold_plots/", replacement = "")) %>%
    mutate(VT_answer = str_replace(VT_answer, pattern = "_VT.png", replacement = "")) %>%
    mutate(correct_VT = case_when(
      VT_answer == VT_textbox2.text ~ "y",
      VT_answer != VT_textbox2.text ~ "n",
      is.na(VT_answer) ~ "n", TRUE ~ as.character(VT_answer))) %>%
    group_by(participant) %>% 
    summarise(VT_no_correct = sum(correct_VT == "y")) %>%
    select("VT_no_correct", "participant")
  
# extract and process monitor and dot pitch information
# we assume standard 16:9 aspect ratio for monitors
  
  monitor_information <- anon_file %>%
    filter(!is.na(height)) %>%
    filter(!is.na(res_width)) %>%
    mutate(res_height = res_width*0.5625,
           width = height*0.5625,
           dot_pitch = ((sqrt(height^2 + width^2))/(sqrt(res_height^2 + res_width^2))) * 25.4) %>%
        select(c("dot_pitch", "participant", "res_width"))
    
  
# extract demographic information
# link slider response numbers to gender categories
  
  demographics <- anon_file %>%
    filter(!is.na(gender_slider.response)) %>%
    mutate(gender_slider.response = recode(gender_slider.response,
                                         `1` = "F",
                                         `2` = "M",
                                         `3` = "NB")) %>%
  select(matches(c("participant",
                          "age_textbox.text",
                          "gender_slider.response")))

# split images column into item and condition columns 

anon_file <- anon_file %>%
  mutate(images = str_replace(images, pattern = "A", replacement = "-A")) %>%
  mutate(images = str_replace(images, pattern = "B", replacement = "-B")) %>%
  mutate(images = str_replace(images, pattern = "C", replacement = "-C")) %>%
  mutate(images = str_replace(images, pattern = "D", replacement = "-D")) %>%
  separate(images, c("item", "size"), sep = "-") %>%
  mutate(size = str_replace(size, pattern = ".png", replacement = "")) %>%
  mutate(item = str_replace(item, pattern = "all_plots/", replacement = ""))

# select relevant columns
# select only experimental items
# add literacy data
# change data types where appropriate
# output this file with suffix 'tidy'

anon_file %>%
  select(c("participant",
                  "item",
                  "size",
                  "slider.response",
                  "my_rs",
                  "total_residuals",
                  "unique_item_no",
                  "session",
                  "trials.thisN")) %>%
  mutate(half = case_when(
    trials.thisN < 93 ~ "First",
    trials.thisN > 92 ~ "Second" )) %>%
  filter(unique_item_no < 181) %>%
  inner_join(literacy, by = "participant") %>%
  inner_join(demographics, by = "participant") %>%
  inner_join(monitor_information, by = "participant") %>%
  inner_join(visual_thresholds, by = "participant") %>%
  mutate(across(matches(c("item", "size")), as_factor)) %>%
  select(-c("__participant")) %>%
  mutate(difference = my_rs - slider.response) %>%
  mutate(size = fct_relevel(size, c('D', 'C', 'B', 'A'))) %>% 
  assign(paste0(unique(anon_file$expName), "_tidy"),
           value = ., envir = .GlobalEnv)
}

# use wrangle function on anonmyised data file 

wrangle(exp3_anon)

# remove anon df from environment

rm(exp3_anon)

# extract age and gender data

extract_age(E3_size_adjustments_tidy)

extract_gender(E3_size_adjustments_tidy)

extract_literacy(E3_size_adjustments_tidy)
```

# Abstract {#abstract-adjusting-size}

Chapter 4 provided strong evidence for the effects of systematically varying the opacities
of scatterplot points on participants' estimates of correlation in positively 
correlated scatterplots. Utilising the same function and experimental paradigm,
I show in a single experiment that systematically varying the sizes of scatterplot
points is able to bias participants' estimates of correlation to a greater degree 
than manipulations that only adjust point opacity. In a condition where point size
decreases non-linearly as a function of residual distance, correlation estimation
is significantly biased upwards to correct for a historic underestimation bias
to a greater degree. I discuss the implications of these findings for the mechanisms
behind both opacity and size adjustments in scatterplots in relation to correlation
estimation, and recommend techniques for those who design with the estimation of
positive correlation in mind.

# Introduction {#introduction-adjusting-size}

While I was successful at changing participants' perceptions of correlation in 
positively correlated scatterplots in Chapter 4, the extent to which these perceptions
were changed was minimal. \ref{fig-estimates-by-r-e2} illustrates how participants'
mean errors in *r* estimation changed as a function of the subjective *r* value.
Scatterplots employing non-linear opacity decay produced the most drastic changes
in correlation estimation, however this was still small, with an effect size of 
Cohen's *d* = 0.19. Recent evidence suggests that with regards to altering percepts
in scatterplots, changes in point size may be more effective than changes in 
opacity. In a fully-reproducible, large sample (N = 150) study, I show that 
systematically altering point size using the same function is not only able to
more effectively correct for the correlation underestimation bias, but is also
able to alter the shape of the correlation estimation curve.

# Related Work {#related-work-adjusting-size}

## Size and Perception 

## Scatterplot Point Size and Correlation Perception

# Methods {#methods-e3}

## Stimuli {#stimuli-e3}

The creation of stimuli in this experiment follows the same general principles
outlined in Section \ref{creating-stimuli}, Chapter 3. As in Chapter 4, equation
1 was used to map point residuals to size values in the two non-linear decay conditions

\begin{equation}
  point_{size/opacity} = 1 - b^{residual}
\end{equation}

## Dot Pitch in Crowdsourced Experiments

## Point Visibility Testing

## Design {#design-e3}

## Procedure {#procedure-e3}

## Participants {#participants-e3}

Normal to corrected-to-normal vision and English fluency were required. Participants who
had completed any of the experiments described in Chapter 4
were prevented from participating. Data were collected from 164 participants.
14 failed more than 2 out of 6 attention check questions,
and, as per the pre-registration, had their submissions rejected from the study.
The data from the remaining 150 participants were included in the full analysis
(`r printnum(E3_size_adjustments_tidy_gender$M)`% male, `r printnum(E3_size_adjustments_tidy_gender$F)`
% female, and `r printnum(E3_size_adjustments_tidy_gender$NB)`% non-binary). 
Participants' mean age was `r printnum(E3_size_adjustments_tidy_age$mean)`
(*SD* = `r printnum(E3_size_adjustments_tidy_age$sd)`). Mean graph literacy score
was `r printnum(E3_size_adjustments_tidy_graph_literacy$mean)` (*SD* =
`r printnum(E3_size_adjustments_tidy_graph_literacy$sd)`). The average
time taken to complete the experiment was 39 minutes (SD = 14 minutes).

# Results {#results-e3}

# Discussion {#discussion-e3}

Increased Correlation Estimation Accuracy

Constant Correlation Estimation Precision

## Training {#training-e3}

## Limitations {#limitations-e3}

## Future Work {#future-work-e3}
