---
title: "conclusion"
output:
  format: 
    latex:
params:
  eval_models: false
  
knitr:
  opts_chunk: 
    cache_comments: false
    crop: true
    
execute: 
  echo: false
  warning: false
  message: false
  include: false     
---

```{r}
#| label: setup

library(tidyverse)
library(MASS)
library(ggdist)
library(ggpubr)
library(conflicted)

source("../shared_functions.R")

labels_all_exp <- c(A_exp1 = "Low Opacity\n(alpha = 0.25)",
                    B_exp1 = "Medium Opacity\n(alpha = 0.5)",
                    C_exp1 = "High Opacity\n(alpha = 0.75)",
                    D_exp1 = "Full Opacity\n(alpha = 1.0)",
                    A_exp2 = "Non-Linear\nOpacity Decay",
                    B_exp2 = "Linear\nOpacity Decay",
                    C_exp2 = "Inverted\nOpacity Decay",
                    D_exp2 = "Full\n Opacity",
                    A_exp3 = "Non-Linear\nSize Decay",
                    B_exp3 = "Linear\nSize Decay",
                    C_exp3 = "Inverted\nSize Decay",
                    D_exp3 = "Standard\nSize",
                    A_exp4 = "Typical Orientation Size\nTypical Orientation Opacity",
                    B_exp4 = "Inverted Orientation Size\nInverted Orientation Opacity",
                    X_exp4 = "Typical Orientation Size\nInverted Orientation Opacity",
                    Y_exp4 = "Inverted Orientation Size\nTypical Orientation Opacity")

# fix conflicts

conflicts_prefer(dplyr::select(), dplyr::filter(), lme4::lmer())
```

Changing the opacities and sizes of points in positively correlated scatterplots
has clear effects on the estimation of correlation. This thesis set out to
investigate whether visual manipulations of scatterplots based on aspects of
human perception might be able to change estimates of correlation, and, in
particular, whether these visual manipulations could be used to correct for a
historical underestimation bias observed since the 1960s.

# Main Findings

Each published experimental chapter in this work builds on the previous and
provides a unique contribution to knowledge about how changing visual features
may affect thoughts and perceptions about correlations.

In \chap{chap:adjusting_opacity}, the opacities of scatterplot points were
altered while participants were tested on a correlation estimation task. The two
published experiments in this chapter were partially based on an earlier pilot
study that ultimately remains unpublished due to data quality issues (see
Section \ref{pilot-study} for a full account of this). In the first experiment in this
chapter, the opacities of scatterplot points were changed in a uniform manner;
this was done with the intent of discovering any general effects of point
opacity on correlation estimation. Evidence was found for an increase in
underestimation error when global point opacity was reduced. This result
provided evidence that, in disagreement with previous work, changing the
opacities of scatterplot points could be used to alter estimates of the
correlation therein. Based on these findings, data from the pilot study, and
evidence that the arrangement of points in scatterplots may be a good proxy for
correlation judgements, Experiment 2 varied point opacity using functions
related to each point's distance from a regression line. In this experiment,
evidence was found that reducing the opacity of a point as it moved further from
the regression line could bias participants' estimates of correlation upwards to
partially correct for the underestimation bias. While this effect was small, it
showed that estimates of correlation could be changed to correct for bias
*without* removing any data from the plot.

\chap{chap:adjusting_size} expanded these findings to a second visual modality
(point size). Building on findings from \chap{chap:adjusting_opacity} and
evidence that the size of a stimulus may have stronger effects on perception
than its opacity, a single experiment took the functions relating opacity and
residual magnitude in the previous chapter and applied them to point size. The
effect on correlation estimation observed here was much stronger than that
observed in \chap{chap:adjusting_opacity}. Additionally, several changes in the
nature of the estimation curve were observed in this experiment. Firstly, while
the shape of the estimation curve when using point opacity manipulations was
identical to that when using no visual adjustments, the use of size decay
manipulations such that point size decreased with residual magnitude resulted in
an estimation curve with a visibly altered shape. Secondly, in previous work,
and in all experiments undertaken in \chap{chap:adjusting_opacity}, precision in
correlation estimation increased as the objective *r* value itself increases;
this behaviour was not observed when employing manipulations that alter point
size. The non-linear decay condition in this experiment remains the most
promising condition with regards to correcting for the underestimation bias.

In a final implementation of the same experimental paradigm,
\chap{chap:interactions_opacity_size} combined point opacity and size
manipulations from the two previous chapters to further explore the potential
for biasing estimates of correlation and the potential mechanisms behind the
effects observed. Using point opacity and size manipulations together in both
congruent (same direction) and incongruent (opposing directions) combinations,
this single experiment showed that employing both manipulations was able to bias
correlation estimates further than either manipulation in isolation, resulting
in a marked overestimation being introduced. This combination is not, however,
linearly additive, adding support to the idea that there are different
mechanisms of action behind the effects of each visual manipulation on
correlation estimation. This experiment provided evidence that, with regards to
using point opacity and size manipulations to bias correlation estimates, there
are few limits. @fig-mean-all-exp summarises the mean estimation errors from all
4 experiments described in \chap{chap:adjusting_opacity},
\chap{chap:adjusting_size}, and \chap{chap:interactions_opacity_size}.

```{r}
#| label: fig-mean-all-exp
#| include: true
#| fig-cap: Estimated Marginal Mean errors in correlation estimation for Experiments 1 to 4. This figure is an amalgam of Figures \ref{fig-e1-estimates}, \ref{fig-e2-estimates}, \ref{fig-e3-estimates}, and \ref{fig-e4-estimates}. 95% confidence intervals are shown as error bars. The vertical dashed line represents no estimation error.
#| fig-asp: 1.5

# grab and merge wrangled csvs together

all_exp_wrangled <- list.files("../data/other_data/",
                               pattern = "^exp.*\\.csv$",
                               full.names = TRUE) %>%
  lapply(read_csv) %>%
  bind_rows() %>%
  mutate(
    exp = sub("^[^_]*_", "", condition),
    condition = fct_relevel(condition,
                            "Y_exp4", "X_exp4", "B_exp4", "A_exp4",
                            "D_exp3", "C_exp3", "B_exp3", "A_exp3",
                            "D_exp2", "C_exp2", "B_exp2", "A_exp2",
                            "D_exp1", "C_exp1", "B_exp1", "A_exp1")
  ) %>%
  mutate(exp = recode(exp,
                      exp1 = "Experiment 1",
                      exp2 = "Experiment 2",
                      exp3 = "Experiment 3",
                      exp4 = "Experiment 4"))

  # base plot

all_exp_wrangled %>%
  mutate(condition = recode(condition, !!!labels_all_exp)) %>%
  ggplot(aes(x = condition, y = emmean*-1, colour = exp)) +
  geom_point(size = 1.75) +
  geom_pointrange(aes(ymin = asymp.LCL*-1, ymax = asymp.UCL*-1)) +
  theme_ggdist() +
  labs(y = "Estimated Marginal Mean of Error",
       x = "Condition") +
  theme(axis.text = element_text(size = 7),
        axis.title = element_text(size = 8),
        legend.position = "none",
        axis.title.y = element_blank()) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  coord_flip() +
  geom_abline(intercept = 0, slope = 0, linetype = 2) +
  scale_color_brewer(palette = "Dark2", aesthetics = "colour") +
  
  # overestimation / underestimation text
  
  annotate(geom = "segment",
                   x = 0.95,
                   xend = 0.95,
                   y = -.02,
                   yend = -.08,
           arrow = arrow(type = "closed", length = unit(0.1, "cm" ))) +
  annotate(geom = "segment",
                   x = 0.95,
                   xend = 0.95,
                   y = .02,
                   yend = .08,
           arrow = arrow(type = "closed", length = unit(0.1, "cm"))) +
  annotate(geom = "text",
           x = 0.7,
           y = -.05,
           label = "Underestimation",
           size = 2.5) +
  annotate(geom = "text",
           x = 0.7,
           y = .05,
           label = "Overestimation",
           size = 2.5) +
  
  # reflected overestimation / underestimation text
  
  annotate(geom = "segment",
                   x = 16,
                   xend = 16,
                   y = -.02,
                   yend = -.08,
           arrow = arrow(type = "closed", length = unit(0.1, "cm" ))) +
  annotate(geom = "segment",
                   x = 16,
                   xend = 16,
                   y = .02,
                   yend = .08,
           arrow = arrow(type = "closed", length = unit(0.1, "cm"))) +
  annotate(geom = "text",
           x = 15.75,
           y = -.05,
           label = "Underestimation",
           size = 2.5) +
  annotate(geom = "text",
           x = 15.75,
           y = .05,
           label = "Overestimation",
           size = 2.5) +
  
  # no estimation error annotations
  
  annotate(geom = "text",
           x = 3.5,
           y = 0.008,
           label = "No Estimation Error",
           size = 2.5,
           angle = 270) +
  annotate(geom = "text",
           x = 8.5,
           y = 0.008,
           label = "No Estimation Error",
           size = 2.5,
           angle = 270) + 
  annotate(geom = "text",
           x = 13.5,
           y = 0.008,
           label = "No Estimation Error",
           size = 2.5,
           angle = 270) +

  # brackets for each experiment
  
  geom_bracket(xmin = 1, xmax = 4, y.position = 0.15, label = "", inherit.aes = F,
               colour = "#E7298AFF") +
  geom_bracket(xmin = 5, xmax = 8, y.position = 0.15, label = "", inherit.aes = F,
               colour = "#7570B3FF") +
  geom_bracket(xmin = 9, xmax = 12, y.position = 0.15, label = "", inherit.aes = F,
               colour = "#D95F02FF") +
  geom_bracket(xmin = 13, xmax = 16, y.position = 0.15, label = "", inherit.aes = F,
               colour = "#1B9E77FF") +

  # labels for each bracket

  annotate("text", x = 2.5, y = 0.1625, label = "Experiment 4", angle = 270, size = 3.25) +
  annotate("text", x = 6.5, y = 0.1625, label = "Experiment 3", angle = 270, size = 3.25) +
  annotate("text", x = 10.5, y = 0.1625, label = "Experiment 2", angle = 270, size = 3.25) +
  annotate("text", x = 14.5, y = 0.1625, label = "Experiment 1", angle = 270, size = 3.25)
```

Extending these findings into a cognitive space, \chap{chap:belief_change}
utilised the congruent typical orientation size and opacity decay condition from
Experiment 4, contextualised as part of news items, to explore whether the
perceptual effects found in the previous three chapters might also be able to
bias people's beliefs about the levels of relatedness between variable pairs. In
this experiment, people's beliefs about a pair of variables were tested before
and after viewing a series of data-identical scatterplots with either the
congruent typical orientation opacity and size decay condition, or with no
opacity or size decay present. While beliefs about relatedness were changed in
each condition, they changed to a greater degree when participants viewed the
scatterplots with opacity and size decay present. This finding provided evidence
that the perceptual effects of systematically altering the opacities and sizes
of scatterplot points can be extended into a cognitive space, and are able to
change what people believe about the levels of relatedness between variable
pairs.

# Relationship to Prior Work

Overall, the work presented in this thesis contributes to wider evidence that
simple changes in visual features can have perceptual \cite{hong_2022} and
cognitive \cite{karduni_2021} effects on those who view such visualisations. The
finding that changing the opacities and sizes of scatterplot points can affect
viewers' estimates of correlation is in direct contradiction to earlier work
\cite{rensink_2012, rensink_2014}, which found no effects of changing point
opacities and sizes in different ways. The novelty of the findings in this
thesis necessarily means that commenting on the potential mechanisms behind them
is difficult. Taking the findings from \chap{chap:adjusting_opacity},
\chap{chap:adjusting_size}, and \chap{chap:interactions_opacity_size} together,
the balance of evidence suggests that decreasing the opacity or size of more
exterior scatterplot points is able to bias estimates of correlation upwards by
reducing the influence of those points when viewers make their estimates. The
results presented in \chap{chap:belief_change} provide evidence that cognitions
about data visualisations can also be altered by small changes in visual
presentation. This is in line with a large body of work demonstrating that small
changes in visual presentation - often of the kind that may be considered only
as aesthetic preference by designers - can lead to large changes in perceptions
and the conclusions that viewers of visualisations draw; see Fygenson et al.,
2024 \cite{fygenson_2024} for examples with bar chart spacing and arrangement,
Moritz et al., 2023 \cite{moritz_2023} for examples with line graphs, and
Bradley et al., 2025 \cite{bradley_2025} for examples with axis limits and
truncation. Franconeri et al. (2021 \cite{franconeri_2021}) and Szafir (2018
\cite{szafir_2018}) both provide excellent wider reviews on how poor design can
lead to visualisation-driven miscommunication and bias.

# Reproducibility

As detailed in \chap{chap:gen_methods}, this project has been conducted
according to the principles of open and reproducible science \cite{ayris_2018}.
Doing so effectively facilitates future replication and authentication of the
published results, in addition to enabling others to learn from where I have
succeeded and failed. Materials are provided throughout for the re-creation of
each individual experimental paper, including Docker containers to reproduce the
specific computational environments used. This thesis also features an
all-in-one Docker implementation, details of which can be found in the
repository associated with this paper.

# Contributions {#contributions-conclusion}

This thesis addresses a long-standing bias in the interpretation of correlation
in positively correlated scatterplots by drawing on influences from perceptual
psychology and data visualisation design. A thorough exploration of the effects
of two salient visual features (opacity and size) on correlation estimation was
carried out over a series of well-powered experiments in
\chap{chap:adjusting_opacity}, \chap{chap:adjusting_size}, and
\chap{chap:interactions_opacity_size}. Furthermore, it was shown in 
\chap{chap:belief_change} that the effects of point opacity and size with regards to the estimation of positive
correlation could be extended into a cognitive space to change peoples' beliefs
about the levels of relatedness between a pair of variables.

The most important contribution this thesis makes is to the visualisation design
and HCI audiences; secondarily, I add to a growing body of work that, taken
together, may one day be able to reveal precisely how viewers' interpret
correlation from scatterplots.

Additionally, I hope to have provided an example of a PhD project conducted from
the ground up in an open and reproducible manner. A PhD project is as much a
story as it is a piece of research, and through acknowledgement of the successes
and failures of the work I have done, along with the employment of a host of
reproducibility-enabling tools, I hope to contribute an honest account of what
one PhD project may look like.

# Implications

## For Design

It is clear from the results presented in this thesis that designers must be
aware of the potential effects on perception and cognition of aesthetic choices
in data visualisation design. This is particularly salient given that changes in
point opacity and size are both common techniques in scatterplot design. The
former is often used to reduce overplotting issues, while the latter is used to
create trivariate bubble charts. Given that many data visualisations are
designed with a range of communicative intentions, designers may be
inadvertently sacrificing effectiveness in one modality (correlation
perception), for clarity in another (reduction in overplotting, for example);
while this trade-off may be acceptable, it is important for designers to be
aware of the effects their designs have on people's thoughts and feelings about
the data stories they are being told.

On the other hand, this work also provides marked opportunities. Where
appropriate, using some form of the point opacity and/or size manipulations
described in this thesis could enhance the communicative efficacy of positively
correlated scatterplots. I would not recommend this until further work has been
done on the potential for misinterpretation from scatterplots such as those
described here (see Section \ref{future-work-chap8}).

## For Society

Given the ubiquity of scatterplots and the fundamental nature of correlation as
a mathematical concept, the findings here have the potential to change the way
that people engage with visualisations depicting relatedness. This is especially
pertinent given the findings detailed in \chap{chap:belief_change} that suggest
that the techniques described here are also able to bias viewers' beliefs about
correlations. Whether or not such techniques would be appropriate is a question
that requires further work to answer.

Additionally, I hope that this project is demonstrative of the need for the
re-examination of common data visualisations from the ground up. In an age where
everything is measured, and the complexity of information increases year on
year, data visualisations are crucial for the communication of difficult and complex topics
to the general public. Many of the most popular, widely-used data visualisations
are not designed to be accurately interpreted by lay people; given that this is
the group that represents the greatest viewership, and that has the most to gain
from engaging with data visualisations, more design must take place that
acknowledges and addresses the ways in which perceptions and cognitions around
data visualisations are biased. Designing in this way requires humility; as
designers, we must recognise that biases exist in common data visualisations and
work thoroughly to address them.

# Limitations

For \chap{chap:adjusting_opacity}, \chap{chap:adjusting_size}, and
\chap{chap:interactions_opacity_size}, the shared experimental method confers
the largest limitation in this thesis. As opposed to previous work that employed
more psychophysical designs, such as correlation discrimination or magnitude
estimation via bisection tasks \cite{rensink_2010, rensink_2017}, Experiments 1
to 4 in this thesis utilised a direct estimation paradigm more akin to much
older correlation perception work
\cite{strahan_1978, bobko_1979, cleveland_1982, lane_1985, lauer_1989, collyer_1990,
meyer_1992}. While this technique supplies robust conclusions for the HCI and
design audiences, it means that making claims about the precise mathematical
relationship between perceived and actual correlation is difficult based on the
data gathered. Similarly, the correlation judgements that participants made in
these experiments must be considered in context; it would not be expected that
people may view 180 data visualisations in sequence during the course of normal,
day-to-day visualisation viewing.

All experiments presented in this thesis were conducted entirely online. Doing
so allowed for the rapid collection of large amounts of high quality data,
however necessarily limits the amount of additional data that can be collected.
A method for determining the dot pitch of participants' monitors was employed
from \chap{chap:adjusting_size} onwards and was included in statistical
modelling paradigms. Generally, this explained little to no variance in
participants' responses. Online experimentation does not allow for the
determination of other factors that may explain variance, such as
head-to-monitor distance or monitor specifications that may influence the visual
quality of scatterplots with altered point opacities or sizes.

# Future Directions {#future-work-chap8}

There are numerous avenues for future work based on this thesis, many of which
are relatively simple extensions of the techniques described. Firstly, all
experiments in this thesis took place using positively correlated scatterplots.
There is evidence that, similarly to how viewers *underestimate* correlation in
positively correlated scatterplots, they may also *overestimate* correlation in
negatively correlated scatterplots \cite{sher_2017}. Uniquely among correlation
visualisations, the visual forms of positively and negatively correlated
scatterplots are symmetrical \cite{harrison_2014}, suggesting that the
techniques modifying point opacity and size described in this thesis may be used
(in a symmetrical manner) to address the overestimation bias. Doing so would
require a relatively simple replication of a subset of the experiments described
here using negatively correlated scatterplots.

This thesis represents a large amount of quantitative data on how viewers
perceive correlation in scatterplots, and how their beliefs about relatedness
may be changed following scatterplot viewing. Before recommendations can be
firmly made about the potential use of these techniques, however, qualitative
work should take place investigating users' thoughts and feelings about novel
visualisations that have been designed based on perceptual metrics. It may be
the case that any potential advantages of the novel scatterplots described
herein are outweighed by distrust or unfamiliarity with novel designs.

Finally, to further investigate precisely *why* the techniques described in this
thesis work, lab-based experiments should be conducted. While more costly and
time-consuming, this would allow for far more control over the precise
experimental set-up, and would additionally allow for the employment of
eye-tracking to understand how changing point opacities and sizes may influence
perceptions of correlation.

# Closing Remarks

The design of data visualisations should be based, at least partially, on the
ways in which those data visualisations are perceived by the people that use
them. The opacities and sizes of points in positively correlated scatterplots
have clear, strong effects on the perceptions of correlation based on them. The
extent to which viewers update their beliefs following scatterplot-viewing can
also depend on the opacities and sizes of scatterplot points. These findings
highlight that small changes in the visual features of common data
visualisations may produce large changes in the perceptions and cognitions of
viewers.
